A New Hybrid Machine Learning Method for Stellar Parameter Inference

### 2025-03-07 paper/paper2/paper.tex
Adjust paper figures

\begin{abstract}
    The advent of machine learning (ML) is revolutionary to numerous scientific disciplines, with a growing number of examples in astronomical spectroscopic inference, as ML is more powerful than traditional techniques.
    Here we introduce a hybrid ML (HML) method combin{ing} automatic differentiation, interpolation, and Bayesian optimization to infer stellar parameters \textbf{of} stellar spectra.
    We study $T_{\mathrm{eff}}$, $\log(g)$, and [Fe/H], but this method \textbf{could be extended} to other \textbf{parameters} such as [$\alpha$/Fe] \textbf{(alpha element abundance)}, C/O \textbf{(carbon-oxygen ratio)}, and $f_{sed}$ \textbf{(sedimentation efficiency)}.
    We first use \texttt{blase} \textbf{to} semi-empirical\textbf{ly recast} spectra \textbf{into} sets of \textbf{Voigt profiles}.
    \texttt{blase} is \textbf{run} on 1,314 \textbf{synthetic} spectra from \textbf{a rectilinear subset of} the PHOENIX model grid \textbf{($T_\text{eff}$: [2300, 12000] K, $\log(g)$: [2, 6] dex, [Fe/H]: [-0.5, 0] dex, $\lambda$: [8038, 12849] \AA)}.
    For \textbf{128,723 detected features}, we map stellar parameters to spectral line parameters using linear interpolation.
    \textbf{This} create\textbf{s} the PHOENIX generator, \textbf{enabling parallelized spectral synthesis}.
    Gaussian Process minimization is used to infer stellar parameters by minimizing \textbf{a} root-mean-square loss \textbf{function}.
    Testing \textbf{210 noise-free models ($T_{\mathrm{eff}}$: [3000, 11000] K, $\log(g)$: [2, 6] dex, [Fe/H]: [-0.5, 0] dex), we find inference errors:} $T_{\mathrm{eff}}$\textbf{: 93 K, $\log(g)$: 0.24 dex, and [Fe/H]: 0.056 dex for $T_\text{eff}$ < 7000 K, and $T_{\mathrm{eff}}$: 347 K, $\log(g)$: 0.26 dex, and [Fe/H]: 0.16 dex for $T_\text{eff} \ge$ 7000 K}.
    \textbf{We also upload online} an archive of \texttt{blase} \textbf{models} of the PHOENIX subset.
    This proof-of-concept \textbf{study} shows that semi-empirical HML is a viable alternative to traditional approaches \textbf{in spectroscopic inference}.
\end{abstract}

@@

\begin{table*}
    \hspace*{0.6cm}\begin{tabular}{llll}
        \hline
        \bf{Parameter} & \bf{Symbol} & \bf{Interval} & \bf{Sampling}\\
        \hline
        Alpha Element Abundance & $\alpha$ & [0] dex & \bf{N/A}\\
        Iron Abundance & [Fe/H] & [-0.5, 0] dex & \bf{0.5 dex}\\
        Effective Temperature & $T_{\mathrm{eff}}$ & [2300, 12000] K & \bf{100 K until 7000 K, then 200 K}\\
        Surface Gravity & $\log(g)$ & [2, 6] dex & \bf{0.5 dex} \\
        Wavelength & $\lambda$ & [8038, 12849] \AA & $\mathbf{R = 500,000}$\\
        \hline
    \end{tabular}
    \caption{The subset of the PHOENIX grid used in this study.
    These limits were imposed to reduce the computational cost of the algorithms and to ensure a rectilinear parameter space in order to work with \texttt{scipy}'s \texttt{RegularGridInterpolator} \citep{scipy}.
    The wavelength limits in particular roughly line up with that of the Habitable Zone Planet Finder (HPF) spectrograph \citep{HPF}.
    This subset is comprised of 1,314 individual spectra: 73 $T_{\mathrm{eff}}$ values, 9 $\log(g)$ values, and 2 [Fe/H] values.}
    \label{tab:table1}
\end{table*}

@@

\textbf{Pre-optimized line centers were required to be equal in order to group lines together.
The values of $\mu$ pre-optimization are recorded by \texttt{blase} to 0.01 \AA (picometer) precision; if they instead used the full available floating point precision, this method would likely detect singular (ungroupable) instances of spectral lines across the grid, completely negating the premise of this study.
Of course, this method can be confused by species that produce spectral lines at very close wavelengths and other confounding factors, likely resulting in some false positives and negatives.
However, we deem it sufficient to approximate this way, as an in-depth treatment would be time-consuming and computationally expensive.}

@@

Examples of the appearance of missing sections in heatmaps where a line does not appear \textbf{can be seen} in \autoref{fig:figure3} and \autoref{fig:figure4}\textbf{, which show heatmaps of the 11617.66 \AA \ C I and the 10123.66 \AA \ He II lines (identified with NIST \citep{NIST}) varying over a solar metallicity slice of the PHOENIX subset stellar parameter space}.
Across the entire PHOENIX subset, \texttt{blase} detected 128,723 spectral lines \textbf{with unique $\mu$}.
\textbf{On average, a PHOENIX spectrum has 9,167 detected spectral lines, with the minimum being 252 ($T_\text{eff}$ = 12000 K, $\log(g)$ = 6 dex, [Fe/H] = -0.5 dex) and the maximum being 34,551 ($T_\text{eff}$ = 2300 K, $\log(g)$ = 3 dex, [Fe/H] = 0 dex).}
Every one of these lines can be visualized as a manifold mapping a 3D stellar parameter vector to a 4D spectral line parameter vector, and every one \textbf{was} interpolated \textbf{to map stellar parameters to spectral line properties}.
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figure3}
    \caption{Heatmap\textbf{s} showing how $\ln(a)$, $\sigma$\textbf{, $\gamma$, and $\mu' - \mu$} vary over the PHOENIX subset slice at solar metallicity of a \textbf{C I} spectral line\textbf{, notable for being detected at nearly all subset grid points}.
    Notice the missing chunk in the top left; \textbf{this could perhaps be due to the line center shifting to the point that \texttt{blase} identified the chunk as a different line entirely (notice the line center offset increasing in magnitude approaching the chunk border).}
    \textbf{Since} \texttt{blase} did not detect a spectral line here, we artificially populate those points with lines that have $\ln(a) = -1000$ \textbf{during interpolation}.}
    \label{fig:figure3}
\end{figure*}
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figure4}
    \caption{Heatmap\textbf{s} showing how $\ln(a)$, $\sigma$\textbf{, $\gamma$, and $\mu' - \mu$} vary over the PHOENIX subset slice at solar metallicity of a He II spectral line.
    \texttt{blase} detects this line at a select chunk of grid points, leading to \textbf{data only existing between $T_\text{eff}$ 3600 K and 6600 K}.}
    \label{fig:figure4}
\end{figure}

@@

Performance results are shown in \autoref{fig:figure7}, and we can see that \textbf{the PHOENIX generator's parallel mode} is much faster than \textbf{serial mode}.
\textbf{While the generator takes around 15 seconds per spectrum in serial mode, this drops to 1 second per spectrum in parallel mode as soon as 30 reconstructions.
The speedup factor between the two methods increases as more spectra are generated, reaching a factor of 20 at 60 reconstructions.
Assuming linear time scaling with wavelength range, according to KORG benchmarks \citep{korg}, which generate spectra over a 500 \AA \ wavelength range, the PHOENIX generator in serial mode is approximately as fast as KORG and $\sim$30 times faster than Turbospectrum \citep{turbospectrum}.
The speedup factor in parallel mode compared to KORG is then the same variable speedup factor from serial mode (20 times faster at 60 reconstructions).}

@@

\begin{figure}[t!]
    \centering
    \includegraphics[width=\columnwidth]{figure7}
    \caption{\textbf{Plots} showing the time taken to reconstruct varying numbers of spectra using the PHOENIX generator (lower is better).
    We can see that the time taken per spectr\textbf{um} for the serial implementation hovers around 15 seconds within run-to-run variance, while the pseudo-parallel implementation continually decreases in time taken per spectrum as the number of inputs increases.
    The speedup factor (higher is better) increases as more spectra are generated, which is also a desirable outcome.}
    \label{fig:figure7}
\end{figure}

@@

We know that in typical observational settings, a coarse estimate for $T_{\mathrm{eff}}$ tends to be fairly well-constrained from ancillary data \textbf{such as photometric color index}.
So when testing the inference algorithm, we limited its search space to only include $T_{\mathrm{eff}}$ which lay within 500 K of the true value on either side.
$\log(g)$ and [Fe/H] were allowed to vary freely.
The test sample consisted of 21 $T_{\mathrm{eff}}$ values\textbf{: [3000, 11000] K in steps of 400 K}, 5 $\log(g)$ values\textbf{: [2, 6] dex in steps of 0.5 dex} , and 2 [Fe/H] values\textbf{: [-0.5, 0] dex in steps of 0.5 dex}, totaling \textbf{210} unique spectra in the test set.
\textbf{Two test passes were run, meaning that 420 inference tests were run in total.}

In \autoref{fig:figure8}, we show the residuals of the inferred stellar parameters with respect to the true values.
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figure8}
    \caption{\textbf{Residual plots for the GP minimization inference algorithm, all plotted with respect to $T_\text{eff}$ (values aggregated across $\log(g)$ and [Fe/H]).
    Notice the interesting change in behavior of all residuals at around $T_\text{eff} = 7000$ K.}}
    \label{fig:figure8}
\end{figure*}
\textbf{We see that there is a distinction between the algorithm's behavior above and below $T_{\mathrm{eff}} = 7000$ K, so we define low and high temperature regimes accordingly.
For $T_{\mathrm{eff}}$ < 7000 K, the inferred} $T_{\mathrm{eff}}$ differed from the true \textbf{value} by an average of \textbf{93} K, $\log(g)$ \textbf{0.24 dex}, and [Fe/H] \textbf{0.056} dex \textbf{(standard deviations of 139 K, 0.40 dex, and 0.15 dex, respectively)}.
\textbf{For $T_{\mathrm{eff}} \ge$ 7000 K, the inferred $T_{\mathrm{eff}}$ differed from the true value by an average of 347 K, $\log(g)$ 0.26 dex, and [Fe/H] 0.16 dex (standard deviations of 293 K, 0.32 dex, and 0.23 dex, respectively).
We see that \texttt{blase3D} is much more effective at predicting stellar parameters of lower temperature spectra, presumably due to the larger amount of spectral lines that can be detected.}

@@

\vspace{0.5cm}
This material is based on work supported by the National Aeronautics and Space Administration under grant No. 80NSSC21K0650 for the NNH20ZDA001N-ADAP:D.2 program. C.V.M. acknowledges support from the Alfred P. Sloan Foundation under grant number FG-2021-16592 and support from the National Science Foundation under grant number 1910969.
\vspace{0.5cm}

@@

\pagebreak
\bibliography{paper}

>>>>

\begin{abstract}
    The advent of machine learning (ML) is revolutionary to numerous scientific disciplines, with a growing number of examples in astronomical spectroscopic inference, as ML is more powerful than traditional techniques.
    Here we introduce a hybrid ML (HML) method combin{ing} automatic differentiation, interpolation, and Bayesian optimization to infer stellar parameters \textbf{of} stellar spectra.
    We study $T_{\mathrm{eff}}$, $\log(g)$, and [Fe/H], but this method \textbf{could be extended} to other \textbf{parameters} such as [$\alpha$/Fe] \textbf{(alpha element abundance)}, C/O \textbf{(carbon-oxygen ratio)}, and $f_{sed}$ \textbf{(sedimentation efficiency)}.
    We first use \texttt{blase} \textbf{to} semi-empirical\textbf{ly recast} spectra \textbf{into} sets of \textbf{Voigt profiles}.
    \texttt{blase} is \textbf{run} on 1,314 \textbf{synthetic} spectra from \textbf{a rectilinear subset of} the PHOENIX model grid \textbf{($T_\text{eff}$: [2300, 12000] K, $\log(g)$: [2, 6], [Fe/H]: [-0.5, 0], $\lambda$: [8038, 12849] \AA)}.
    For \textbf{128,723 detected features}, we map stellar parameters to spectral line parameters using linear interpolation.
    \textbf{This} create\textbf{s} the PHOENIX generator, \textbf{enabling parallelized spectral synthesis}.
    Gaussian Process minimization is used to infer stellar parameters by minimizing \textbf{a} root-mean-square loss \textbf{function}.
    Testing \textbf{210 noise-free models ($T_{\mathrm{eff}}$: [3000, 11000] K, $\log(g)$: [2, 6], [Fe/H]: [-0.5, 0]), we find inference errors:} $T_{\mathrm{eff}}$\textbf{: 93 K, $\log(g)$: 0.24, and [Fe/H]: 0.056 for $T_\text{eff}$ < 7000 K, and $T_{\mathrm{eff}}$: 347 K, $\log(g)$: 0.26, and [Fe/H]: 0.16 for $T_\text{eff} \ge$ 7000 K}.
    \textbf{We also upload online} an archive of \texttt{blase} \textbf{models} of the PHOENIX subset.
    This proof-of-concept \textbf{study} shows that semi-empirical HML is a viable alternative to traditional approaches \textbf{in spectroscopic inference}.
\end{abstract}

@@

\begin{table*}
    \hspace*{0.6cm}\begin{tabular}{llll}
        \hline
        \bf{Parameter} & \bf{Symbol} & \bf{Interval} & \bf{Sampling}\\
        \hline
        Alpha Element Abundance & $\alpha$ & [0]& \bf{N/A}\\
        Iron Abundance & [Fe/H] & [-0.5, 0] & \bf{0.5}\\
        Effective Temperature & $T_{\mathrm{eff}}$ & [2300, 12000] K & \bf{100 K until 7000 K, then 200 K}\\
        Surface Gravity & $\log(g)$ & [2, 6] & \bf{0.5} \\
        Wavelength & $\lambda$ & [8038, 12849] \AA & $\mathbf{R = 500,000}$\\
        \hline
    \end{tabular}
    \caption{The subset of the PHOENIX grid used in this study.
    These limits were imposed to reduce the computational cost of the algorithms and to ensure a rectilinear parameter space in order to work with \texttt{scipy}'s \texttt{RegularGridInterpolator} \citep{scipy}.
    The wavelength limits in particular roughly line up with that of the Habitable Zone Planet Finder (HPF) spectrograph \citep{HPF}.
    This subset is comprised of 1,314 individual spectra: 73 $T_{\mathrm{eff}}$ values, 9 $\log(g)$ values, and 2 [Fe/H] values.}
    \label{tab:table1}
\end{table*}

@@

\textbf{Pre-optimized line centers were required to be equal in order to group lines together.
The values of $\mu$ pre-optimization are recorded by \texttt{blase} to 0.01 \AA \ (picometer) precision; if they instead used the full available floating point precision, this method would likely detect singular (ungroupable) instances of spectral lines across the grid, completely negating the premise of this study.
Of course, this method can be confused by species that produce spectral lines at very close wavelengths and other confounding factors, likely resulting in some false positives and negatives.
However, we deem it sufficient to approximate this way, as an in-depth treatment would be time-consuming and computationally expensive.}

@@

Examples of the appearance of missing sections in heatmaps where a line does not appear \textbf{can be seen} in \autoref{fig:figure3} and \autoref{fig:figure4}\textbf{, which show heatmaps of the 11617.66 \AA \ C I and the 10123.66 \AA \ He II lines (identified with NIST \citep{NIST}) varying over a solar metallicity slice of the PHOENIX subset stellar parameter space}.
Across the entire PHOENIX subset, \texttt{blase} detected 128,723 spectral lines \textbf{with unique $\mu$}.
\textbf{On average, a PHOENIX spectrum has 9,167 detected spectral lines, with the minimum being 252 ($T_\text{eff}$ = 12000 K, $\log(g)$ = 6, [Fe/H] = -0.5) and the maximum being 34,551 ($T_\text{eff}$ = 2300 K, $\log(g)$ = 3, [Fe/H] = 0).}
Every one of these lines can be visualized as a manifold mapping a 3D stellar parameter vector to a 4D spectral line parameter vector, and every one \textbf{was} interpolated \textbf{to map stellar parameters to spectral line properties}.
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figure3}
    \caption{Heatmap\textbf{s} showing how $\ln(a)$, $\sigma$\textbf{, $\gamma$, and $\mu' - \mu$} vary over the PHOENIX subset slice at solar metallicity of a \textbf{C I} spectral line\textbf{, notable for being detected at nearly all subset grid points}.
    Notice the missing chunk in the top left; \textbf{this could perhaps be due to the line center shifting to the point that \texttt{blase} identified the chunk as a different line entirely (notice the line center offset increasing in magnitude approaching the chunk border).}
    \textbf{Since} \texttt{blase} did not detect a spectral line here, we artificially populate those points with lines that have $\ln(a) = -1000$ \textbf{during interpolation}.}
    \label{fig:figure3}
\end{figure*}
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figure4}
    \caption{Heatmap\textbf{s} showing how $\ln(a)$, $\sigma$\textbf{, $\gamma$, and $\mu' - \mu$} vary over the PHOENIX subset slice at solar metallicity of a He II spectral line.
    \texttt{blase} detects this line at a select chunk of grid points, leading to \textbf{data only existing between $T_\text{eff}$ 3600 K and 6600 K}.}
    \label{fig:figure4}
\end{figure}

@@

Performance results are shown in \autoref{fig:figure7}, and we can see that \textbf{the PHOENIX generator's parallel mode} is much faster than \textbf{serial mode}.
\begin{figure}[t!]
    \centering
    \includegraphics[width=\columnwidth]{figure7}
    \caption{\textbf{Plots} showing the time taken to reconstruct varying numbers of spectra using the PHOENIX generator (lower is better).
    We can see that the time taken per spectr\textbf{um} for the serial implementation hovers around 15 seconds within run-to-run variance, while the pseudo-parallel implementation continually decreases in time taken per spectrum as the number of inputs increases.
    The speedup factor (higher is better) increases as more spectra are generated, which is also a desirable outcome.}
    \label{fig:figure7}
\end{figure}
\textbf{While the generator takes around 15 seconds per spectrum in serial mode, this drops to 1 second per spectrum in parallel mode as soon as 30 reconstructions.
The speedup factor between the two methods increases as more spectra are generated, reaching a factor of 20 at 60 reconstructions.
Assuming linear time scaling with wavelength range, according to KORG benchmarks \citep{korg}, which generate spectra over a 500 \AA \ wavelength range, the PHOENIX generator in serial mode is approximately as fast as KORG and $\sim$30 times faster than Turbospectrum \citep{turbospectrum}.
The speedup factor in parallel mode compared to KORG is then the same variable speedup factor from serial mode (20 times faster at 60 reconstructions).}

@@

@@

We know that in typical observational settings, a coarse estimate for $T_{\mathrm{eff}}$ tends to be fairly well-constrained from ancillary data \textbf{such as photometric color index}.
So when testing the inference algorithm, we limited its search space to only include $T_{\mathrm{eff}}$ which lay within 500 K of the true value on either side.
$\log(g)$ and [Fe/H] were allowed to vary freely.
The test sample consisted of 21 $T_{\mathrm{eff}}$ values\textbf{: [3000, 11000] K in steps of 400 K}, 5 $\log(g)$ values\textbf{: [2, 6] in steps of 0.5} , and 2 [Fe/H] values\textbf{: [-0.5, 0] in steps of 0.5}, totaling \textbf{210} unique spectra in the test set.
\textbf{Two test passes were run, meaning that 420 inference tests were run in total.}

In \autoref{fig:figure8}, we show the residuals of the inferred stellar parameters with respect to the true values.
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figure8}
    \caption{\textbf{Residual plots for the GP minimization inference algorithm, all plotted with respect to $T_\text{eff}$ (values aggregated across $\log(g)$ and [Fe/H]).
    Notice the interesting change in behavior of all residuals at around $T_\text{eff} = 7000$ K.}}
    \label{fig:figure8}
\end{figure*}
\textbf{We see that there is a distinction between the algorithm's behavior above and below $T_{\mathrm{eff}} = 7000$ K, so we define low and high temperature regimes accordingly.
It is interesting that this observed change occurs where the PHOENIX grid changes in $T_\text{eff}$ sampling from 100 K to 200 K.}

\textbf{For $T_{\mathrm{eff}}$ < 7000 K, the inferred} $T_{\mathrm{eff}}$ differed from the true \textbf{value} by an average of \textbf{93} K, $\log(g)$ \textbf{0.24}, and [Fe/H] \textbf{0.056} \textbf{(standard deviations of 139 K, 0.40, and 0.15, respectively)}.
\textbf{For $T_{\mathrm{eff}} \ge$ 7000 K, the inferred $T_{\mathrm{eff}}$ differed from the true value by an average of 347 K, $\log(g)$ 0.26, and [Fe/H] 0.16 (standard deviations of 293 K, 0.32, and 0.23, respectively).
We see that \texttt{blase3D} is more effective and generally more self-consistent when predicting stellar parameters of lower temperature spectra.}

@@

\textbf{All figures in this paper are publicly available on GitHub at \url{https://github.com/Sujay-Shankar/blase3D} under \texttt{paper/paper2}.}

\section*{}
This material is based on work supported by the National Aeronautics and Space Administration under grant No. 80NSSC21K0650 for the NNH20ZDA001N-ADAP:D.2 program. C.V.M. acknowledges support from the Alfred P. Sloan Foundation under grant number FG-2021-16592 and support from the National Science Foundation under grant number 1910969.
\vspace*{0.5cm}

@@

\bibliography{paper}
